Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and attentional function: A cross-sectional study based on the parametric g-formula

### 2024-07-24 output/paper/manuscript.qmd
Added modified title page.
Comments from MV before submission.

Taken together, these results suggest that the negative influence of exposure to certain {{< acr edcs >}} on neurodevelopmental outcomes might be mediated, at least partially, by disruption of the {{< acr hpa >}} axis' homeostasis.
In the present study, we thus estimated cross-sectional associations between 1) non-persistent {{< acr edcs >}} and attentional function, 2) non-persistent {{< acr edcs >}} and glucocorticosteroids, and 3) glucocorticosteroids and attentional function, using the parametric g-formula and {{< acr mcs first_use=true >}}, in children of a large network of cohorts in Europe.

@@

Cognitive and motor function outcomes were assessed with standardized, non-linguistic, and culturally blind computer tests, including the {{< acr ant first_use=true >}} [@RuedaFanMcCandliss:2004], which provides a measure of efficiency of attentional function. The tests were administered in a standardized way, and with minimal interference from the field workers. Further information can be found elsewhere [@RuedaFanMcCandliss:2004; @FornsEsnaolaLopez-Vicente:2014; @MaitreBontCasas:2018]. The outcome of interest for the present study is the {{< acr hitrtse first_use=true >}} [@SunyerEsnaolaAlvarez-Pedrerol:2015], a measure of response speed consistency throughout the test. A high {{< acr hitrtse >}} indicates highly variable reaction times, and is considered a measure of inattentiveness.

@@

For each research question, defined by a specific type of exposure and outcome, the minimal set of covariates for inclusion in the analyses was selected on the basis of a {{< acr dag >}} built with `DAGitty` [@TextorvanderZanderGilthorpe:2016] and `ggdag` [@Barrett:2023]. The sets of covariates were selected to estimate the total effect of the exposure on the outcome. For effect estimation of the {{< acr edcs >}} on glucocorticosteroids and of glucocorticosteroids on {{< acr hitrtse >}}, these sets were also sufficient to estimate direct effects.
Sample-specific creatinine values were included in the models to adjust for possible dilution effects.
Further, each minimal adjustment set was *augmented* with precision covariates, defined as the set of parents variable of the outcome that are not parents of the exposure. Common confounders were cohort, ethnicity, sex, age, height, weight, and head circumference of the child, consumption of fish, fruit, vegetables, organic food, and fast food with a food frequency questionnaire, maternal tobacco consumption, family financial situation and affluence scale (FAS). Models for estimating the effects of {{< acr edcs >}} on {{< acr hitrtse >}} were further adjusted for child breastfeeding, prenatal maternal active and passive smoking, urine creatinine, child mood and rest before assessment, child neuropsychological diagnosis, marital status, season, and fasting time before assessment. Models for estimating the effects of {{< acr edcs >}} on glucocorticosteroids were further adjusted for urine creatinine, season, and fasting time before assessment. Models for estimating the effects of glucocorticosteroids on {{< acr hitrtse >}} were further adjusted for child breastfeeding, prenatal maternal active and passive smoking, marital status, {{< acr edcs >}}, urine creatinine, child mood and rest before assessment, and child neuropsychological diagnosis. The adjustment sets are provided in the Supplementary Material as text files compatible with DAGitty. A simplified {{< acr swig >}} is provided in @fig-dag. Codebooks for the used covariates are provided in @supptbl-codebooks.

@@

Concentrations of the glucocorticosteroids were classified as quantifiable, below the {{< acr loq >}}, possible interference or out of range, and not detected. For each metabolite, we computed the fraction of values below the {{< acr loq >}} and not detected, both within each cohort and overall. We proceeded to impute these values using half the value of the corresponding {{< acr loq >}}, for those metabolites that had less than `r threshold_within_lod`% of missings within each cohort and `r threshold_overall_lod`% of missings overall. Information about the {{< acr lloq >}} for the glucocorticosteroids is provided in @supptbl-lloq-mets. The remaining missing values were imputed using kNN from the `VIM` R package [@KowarikTempl:2016], for those metabolites that had less than `r threshold_within_nan`% of remaining missings within each cohort and `r threshold_overall_nan`% of remaining missings overall. We used `r knn_k` nearest neighbors. We natural log-transformed them to improve model fit, assessed with posterior predictive checks. To do so, replicated data were simulated with the fitted models and compared to the observed data. We used the `check_predictions` function from the `performance` R package using the default arguments [@LudeckeBen-ShacharPatil:2021]. Values of total cortisol, cortisone, and corticosterone production were expressed in nanograms per millilitre (ng/ml).

Concentrations of the non-persistent {{< acr edcs >}} were classified as quantifiable, below the {{< acr lod >}}, possible interference or out of range, and not analysed. Concentrations below the {{< acr lod >}} were singly imputed using a quantile regression approach for the imputation of left-censored missing data, as implemented in the `impute.QRILC` function from the `imputeLCMD` R package [@lazar2015imputelcmd]. Information about the lower limits of detection can be found in @HaugSakhiCequier:2018. Chemicals with more than 70% of observations below the {{< acr lod >}} were excluded from the present study. Remaining missing values were imputed similarly using kNN. Values of the chemicals were expressed in $\mu$grams per litre ($\mu$g/L).

@@

We estimated {{< acr mcs >}} with the parametric g-formula, a method of standardization. The parametric g-formula involves the following steps: 1) fit a outcome model including both covariates and balancing weights; 2) create two new datasets identical to the original one but with the exposure shifted according to a user-specified dynamic intervention set by a deterministic function of the observed exposure levels; 3) use the outcome model to compute adjusted predictions in the two counterfactual datasets; 4) compute the difference between the means of the adjusted predictions in the counterfactual datasets. The causal parameter of interest was thus specified as the difference in the expected counterfactual outcomes under the shifted exposure levels $\left( \mathbb{E} \left[ Y^{d_1} \right] - \mathbb{E} \left[ Y^{d_2} \right] \right)$. In order for this parameter to be identified, the usual causal identifiability conditions (no unmeasured confounding, positivity, and consistency) are required. Since these conditions are likely not satisfied, we focused on the estimation of a statistical estimand that is as close as possible to the causal parameter of interest.

@@

The R code to reproduce analyses and results is available online (https://github.com/lorenzoFabbri/paper-helixSC-neuro).

@@

@Fig-marginal-1 presents the forest plot for the {{< acr mcs >}} on the logarithmic scale of the non-persistent {{< acr edcs >}} on {{< acr hitrtse >}}. For most {{< acr edcs >}}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the values of {{< acr hitrtse >}} and thus lower attention. Most of the {{< acr cis >}} included the null effect, though. Noteworthy effects were observed for the paraben MEPA (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`). The {{< acr opp >}} DETP was negatively associated with {{< acr hitrtse >}} (`r format_res(meffects1, "detp", "HRT-SE")`).

@Fig-marginal-2 presents the forest plot for the {{< acr mcs >}} on the logarithmic scale of the non-persistent {{< acr edcs >}} on total cortisone, cortisol, and corticosterone production. For most {{< acr edcs >}}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the total production of these metabolites. Exceptions were BUPA, which was associated with negative {{< acr mcs >}} for all three outcomes, and MiBP, which was associated with a negative {{< acr mc >}} for total cortisone production only.
The majority of the effects for the phenols and phthalate metabolites included the null. The phenol BPA showed the largest {{< acr mcs >}} across all three outcomes (cortisone production, `r format_res(meffects2, "bpa", "cortisone production")`; cortisol production, `r format_res(meffects2, "bpa", "cortisol production")`; corticosterone production, `r format_res(meffects2, "bpa", "corticosterone production")`).

@@

In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to certain non-persistent {{< acr edcs >}} was associated with attentional function (MEPA, MEHP, oh-MiNP, and oxo-MiNP), and with total production of cortisol, cortisone, and corticosterone (DEP, DMP, DMTP, BPA, ETPA, MEPA, MEHP, oh-MiNP, and oxo-MiNP). Increased production of these glucocorticosteroids did not seem to affect attentional function. Some of these associations differed for females and males.

@@

In children aged 6 to 11 years, higher levels of {{< acr dap >}} metabolites were associated with lower scores of {{< acr iq >}} and verbal comprehension, especially in boys [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015], while higher levels of diethylphosphate metabolites were associated with lower working memory scores [@CartierWarembourgLeManer-Idrissi:2016]. There is also preliminary evidence of a possible association between exposure to certain {{< acr opps >}} and {{< acr adhd >}} in children [@BouchardBellingerWright:2010; @YuDuChiou:2016]. In our study we found a negative association between DETP and attentional function, indicating lower values of {{< acr hitrtse >}} with higher levels of DETP, with concordant results in the sex-stratified analysis. While we cannot exclude a protective effect of DETP in relation to attentional function, we note that in our study DETP has a higher percentage of missing values compared to the other {{< acr opps >}} and that other imputation strategies might lead to different associations.

Preliminary evidence is also available for several phthalate metabolites in relation to cognitive development in childhood. Higher levels of di(2-ethylhexyl) phthalate metabolites (including MEHP, MEHHP, and MEOHP) were associated with lower intelligence scores in children aged 2 to 12 years [@HuangChenSu:2015], lower scores of {{< acr iq >}} and verbal intelligence, more omission errors (a measure of inattention), and higher scores of response time variability (a measure of sustained attention) in 6-year old Korean children [@KimHongShin:2017], poorer fine motor skills in preadolescent boys [@BalalianWhyattLiu:2019], and lower intelligence scores in 7-year old children [@VilmandBeckBilenberg:2023]. Further associations were found for higher levels of MEOHP with lower scores of {{< acr iq >}} [@HuangChenSu:2015] and verbal intelligence in Taiwanese children aged 6 to 12 years [@HuangTsaiChen:2017], and for higher levels of dibutyl phthalate metabolites (MnBP and MiBP) with impaired verbal intelligence [@HuangTsaiChen:2017].
Few studies have looked into different classes of non-persistent {{< acr edcs >}}. Shoaff et al. investigated cross-sectional associations between multiple {{< acr edcs >}} and {{< acr adhd >}}-related behaviors in 15-year old adolescents, finding a higher risk of {{< acr adhd >}}-related behavior problems with higher levels of antiandrogenic phthalate metabolites (molar sum of MnBP, MiBP, MBzP, MEHP, MEHHP, MEOHP, MECPP, monocarboxyoctyl phthalate, monohydroxyisobutyl phthalate (MHiBP), monohydroxybutyl phthalate (MHBP), and mono-isononyl phthalate) and the molar sum of di(2-ethylhexyl) phthalate (DEHP) metabolites (MECPP, MEHHP, MEOHP, and MEHP), especially in boys [@ShoaffCoullWeuve:2020]. Our findings, indicating that short-term childhood exposure to certain phthalate metabolites (MEHP, oh-MiNP, and oxo-MiNP) was associated with attentional function, adds to this growing evidence base suggesting that childhood phthalate exposure may impact child neurodevelopment.

Among phenols, some studies provide preliminary evidence of an association between BPA and {{< acr adhd >}} in children aged 8 to 15 years [@TewarAuingerBraun:2016] and in a case-control study of children aged 6 to 12 years [@LiZhangKuang:2018], especially in boys. Except for working memory, there was no evidence of an association between BPA and cognitive abilities in Spanish boys aged 9 to 11 years [@Rodriguez-CarrilloMustielesPerez-Lobato:2019]. We did not observe an association between BPA and attention function in the present study, but this study is the first to suggest that childhood exposure to MEPA may be associated with attentional function.

@@

We are also not aware of prior epidemiological studies specifically investigating the effects of elevated levels of glucocorticosteroids in relation to attentional function, although there is evidence that under- or over-production of glucocorticosteroids interfere with the normal development of the brain, possibly in a sex-specific manner [@LupienMcEwenGunnar:2009].

Our findings should be interpreted in light of the following strengths and limitations. Strengths include its relative large sample size and its inclusion of multiple classes of non-persistent {{< acr edcs >}}. Further, this study used pooled urine samples for chemical assessment to reduce measurement error, since it is known that these specific {{< acr edcs >}} have very short half-lives [@PerrierGiorgis-AllemandSlama:2016; @CasasBasaganaSakhi:2018]. We decided to model both the *treatment* mechanisms, for the estimation of balancing weights, and the outcomes, with traditional covariates adjustment, to obtain *doubly robust* effect estimates. Finally, we decided not to interpret our results by focusing on the estimated coefficients of possibly misspecified regression models, but by making use of the g-computation procedure.

Limitations include the cross-sectional design of the present study. Importantly, the non-persistent {{< acr edcs >}} were measured in a pool of night and morning urine samples before the clinical visit, to represent exposure over the previous day, whereas the glucocorticosteroids were measured in the night urine sample. Cross-sectional designs also have limitations to study the etiologic period of susceptibility of attentional function. Prior studies have investigated associations between short-term effects of air pollution and attentional function, with null results in adolescents but higher response times in adults [@GignacBarrera-GomezPersavento:2021; @GignacRighiToran:2022]. Finally, longitudinal studies are necessary for a formal causal mediation analysis.
Although we included a wide range of confounders, there is the possibility, as with other observational studies, of residual confounding, which might lead to a bias away from the null. We decided not to include prenatal levels of the {{< acr edcs >}} as confounders due to the *small* correlation between prenatal and childhood levels of these non-persistent chemicals [@HaugSakhiCequier:2018].
There is further the possibility of misspecification of the outcome model, although we included a spline of the exposure to relax some of the linearity assumptions. Finally, we acknowledge that the present study does not take into account the possible mixture effect of these chemicals. While methods for causal mixture analysis are available [@KeilBuckleyO:2020], we also emphasize that using these tools without thoughtful consideration may lead to misleading results [@WebsterWeisskopf:2020].

>>>>

Taken together, these results suggest that disruption of the {{< acr hpa >}} axis' homeostasis may be one biological mechanism by which {{< acr edcs >}} may affect neurodevelopment.

In the present study, we estimated cross-sectional associations between 1) non-persistent {{< acr edcs >}} and attentional function, 2) non-persistent {{< acr edcs >}} and glucocorticosteroids, and 3) glucocorticosteroids and attentional function in children of a multi-country cohort in Europe. We used the parametric g-formula, a causal inference technique, and {{< acr mcs first_use=true >}} to obtain robust effect estimates and overcome limitations of classical regression models.

@@

Cognitive and motor function outcomes were assessed with standardized, non-linguistic, and culturally blind computer tests, including the {{< acr ant first_use=true >}} [@RuedaFanMcCandliss:2004], which provides a measure of efficiency of attentional function. The tests were administered in a standardized way, and with minimal interference from the field workers. Further information can be found elsewhere [@RuedaFanMcCandliss:2004; @FornsEsnaolaLopez-Vicente:2014; @MaitreBontCasas:2018]. The outcome of interest for the present study is the {{< acr hitrtse first_use=true >}} [@SunyerEsnaolaAlvarez-Pedrerol:2015], a measure of response speed consistency throughout the test. A high {{< acr hitrtse >}} indicates highly variable reaction times and is considered a measure of inattentiveness.

@@

For each research question, defined by a specific type of exposure and outcome, the minimal set of covariates for inclusion in the analyses was selected on the basis of a {{< acr dag >}} built with `DAGitty` [@TextorvanderZanderGilthorpe:2016] and `ggdag` [@Barrett:2023]. The sets of covariates were selected to estimate the total effect of the exposure on the outcome. For effect estimation of the {{< acr edcs >}} on glucocorticosteroids and of glucocorticosteroids on {{< acr hitrtse >}}, these sets were also sufficient to estimate direct effects.
Sample-specific creatinine values were included in the models to adjust for possible dilution effects.
Further, each minimal adjustment set was *augmented* with precision covariates, defined as the set of parent variables of the outcome that are not parents of the exposure. Common confounders were cohort, ethnicity, sex, age, height, weight, and head circumference of the child, consumption of fish, fruit, vegetables, organic food, and fast food with a food frequency questionnaire, maternal tobacco consumption, family financial situation and affluence scale (FAS). Models for estimating the effects of {{< acr edcs >}} on {{< acr hitrtse >}} were further adjusted for child breastfeeding, prenatal maternal active and passive smoking, urine creatinine, child mood and rest before assessment, child neuropsychological diagnosis, maternal marital status, season, and fasting time before assessment. Models for estimating the effects of {{< acr edcs >}} on glucocorticosteroids were further adjusted for urine creatinine, season, and fasting time before assessment. Models for estimating the effects of glucocorticosteroids on {{< acr hitrtse >}} were further adjusted for child breastfeeding, prenatal maternal active and passive smoking, marital status, {{< acr edcs >}}, urine creatinine, child mood and rest before assessment, and child neuropsychological diagnosis. The adjustment sets are provided in the Supplementary Material as text files compatible with DAGitty. A simplified {{< acr swig >}} is provided in @fig-dag. Codebooks for the used covariates are provided in @supptbl-codebooks.

@@

Concentrations of the glucocorticosteroids were classified as quantifiable, below the {{< acr loq >}}, possible interference or out of range, and not detected. For each metabolite, we computed the fraction of values below the {{< acr loq >}} and not detected, both within each cohort and overall. We proceeded to impute these values using half the value of the corresponding {{< acr loq >}}, for those metabolites that had less than `r threshold_within_lod`% of non-detected within each cohort and `r threshold_overall_lod`% of non-detected overall. Information about the {{< acr lloq >}} for the glucocorticosteroids is provided in @supptbl-lloq-mets. Missing values were imputed using kNN from the `VIM` R package [@KowarikTempl:2016], for those metabolites that had less than `r threshold_within_nan`% of missings within each cohort and `r threshold_overall_nan`% of missings overall. We used `r knn_k` nearest neighbors. We natural log-transformed them to improve model fit, assessed with posterior predictive checks. To do so, replicated data were simulated with the fitted models and compared to the observed data. We used the `check_predictions` function from the `performance` R package using the default arguments [@LudeckeBen-ShacharPatil:2021]. Values of total cortisol, cortisone, and corticosterone production were expressed in nanograms per millilitre (ng/ml).

Concentrations of the non-persistent {{< acr edcs >}} were classified as quantifiable, below the {{< acr lod >}}, possible interference or out of range, and not analysed. Concentrations below the {{< acr lod >}} were singly imputed using a quantile regression approach for the imputation of left-censored missing data, as implemented in the `impute.QRILC` function from the `imputeLCMD` R package [@lazar2015imputelcmd]. Information about the lower limits of detection can be found in @HaugSakhiCequier:2018. Chemicals with more than 70% of observations below the {{< acr lod >}} were excluded from the present study. Missing values were imputed similarly using kNN. Values of the chemicals were expressed in $\mu$grams per litre ($\mu$g/L).

@@

We estimated {{< acr mcs >}} with the parametric g-formula, a method of standardization. The parametric g-formula involves the following steps: 1) fit a outcome model including both covariates and balancing weights; 2) create two new datasets identical to the original one but with the exposure shifted according to a user-specified dynamic intervention set by a deterministic function of the observed exposure levels; 3) use the outcome model to compute adjusted predictions in the two counterfactual datasets; 4) compute the difference between the means of the adjusted predictions in the counterfactual datasets. The causal parameter of interest was thus specified as the difference in the expected counterfactual outcomes under the shifted exposure levels $\left( \mathbb{E} \left[ Y^{d_1} \right] - \mathbb{E} \left[ Y^{d_2} \right] \right)$. For this parameter to be identified, the usual causal identifiability conditions (no unmeasured confounding, positivity, and consistency) are required. Since these conditions are likely not satisfied, we focused on the estimation of a statistical estimand that is as close as possible to the causal parameter of interest.

@@

@@

@Fig-marginal-1 presents the forest plot for the {{< acr mcs >}} on the logarithmic scale of the non-persistent {{< acr edcs >}} on {{< acr hitrtse >}}. For most {{< acr edcs >}}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the values of {{< acr hitrtse >}} and thus lower attention. Most of the {{< acr cis >}} included the null effect, though. Noteworthy associations were observed for the paraben MEPA (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`). The {{< acr opp >}} DETP was associated with lower {{< acr hitrtse >}} (`r format_res(meffects1, "detp", "HRT-SE")`).

@Fig-marginal-2 presents the forest plot for the {{< acr mcs >}} on the logarithmic scale of the non-persistent {{< acr edcs >}} on total cortisone, cortisol, and corticosterone production. For most {{< acr edcs >}}, a cohort-specific increase in the levels of the exposures from the 10th to the 90th percentiles was associated with a positive {{< acr mc >}}, indicating an increase in the total production of these metabolites. Exceptions were BUPA, which was associated with negative {{< acr mcs >}} for all three outcomes, and MiBP, which was associated with a negative {{< acr mc >}} for total cortisone production only.
Most of the effects for the phenols and phthalate metabolites included the null. The phenol BPA showed the largest {{< acr mcs >}} across all three outcomes (cortisone production, `r format_res(meffects2, "bpa", "cortisone production")`; cortisol production, `r format_res(meffects2, "bpa", "cortisol production")`; corticosterone production, `r format_res(meffects2, "bpa", "corticosterone production")`).

@@

In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to certain non-persistent {{< acr edcs >}} was associated with lower attentional function (MEPA, MEHP, oh-MiNP, and oxo-MiNP), and with an increase in total production of cortisol, cortisone, and corticosterone (DEP, DMP, DMTP, BPA, ETPA, MEPA, MEHP, oh-MiNP, and oxo-MiNP). Some of these associations differed for females and males. Increased production of these glucocorticosteroids was not associated with attentional function in the study population overall, but showed associations with lower attentional function in males and higher attentional function in females.

@@

In children aged 6 to 11 years, higher levels of {{< acr dap >}} metabolites were associated with lower scores of {{< acr iq >}} and verbal comprehension, especially in boys [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015], while higher levels of diethylphosphate metabolites were associated with lower working memory scores [@CartierWarembourgLeManer-Idrissi:2016]. There is also preliminary evidence of a possible association between exposure to certain {{< acr opps >}} and {{< acr adhd >}} in children [@BouchardBellingerWright:2010; @YuDuChiou:2016]. In our study we found higher levels of DETP to be associated with better attentional function, with concordant results in the sex-stratified analysis. While we cannot exclude a protective effect of DETP in relation to attentional function, we note that in our study DETP has a higher percentage of missing values compared to the other {{< acr opps >}} and that other imputation strategies might lead to different associations.

Previous evidence is also available for several phthalate metabolites in relation to cognitive development in childhood. Higher levels of di(2-ethylhexyl) phthalate metabolites (including MEHP, MEHHP, and MEOHP) were associated with lower intelligence scores in children aged 2 to 12 years [@HuangChenSu:2015], lower scores of {{< acr iq >}} and verbal intelligence, more omission errors (a measure of inattention), and higher scores of response time variability (a measure of sustained attention) in 6-year old Korean children [@KimHongShin:2017], poorer fine motor skills in preadolescent boys [@BalalianWhyattLiu:2019], and lower intelligence scores in 7-year old children [@VilmandBeckBilenberg:2023]. Further associations were found for higher levels of MEOHP with lower scores of {{< acr iq >}} [@HuangChenSu:2015] and verbal intelligence in Taiwanese children aged 6 to 12 years [@HuangTsaiChen:2017], and for higher levels of dibutyl phthalate metabolites (MnBP and MiBP) with impaired verbal intelligence [@HuangTsaiChen:2017].
Few studies have investigated different classes of non-persistent {{< acr edcs >}}. Shoaff et al. investigated cross-sectional associations between multiple {{< acr edcs >}} and {{< acr adhd >}}-related behaviors in 15-year old adolescents, finding a higher risk of {{< acr adhd >}}-related behavior problems with higher levels of antiandrogenic phthalate metabolites (molar sum of MnBP, MiBP, MBzP, MEHP, MEHHP, MEOHP, MECPP, monocarboxyoctyl phthalate, monohydroxyisobutyl phthalate (MHiBP), monohydroxybutyl phthalate (MHBP), and mono-isononyl phthalate) and the molar sum of di(2-ethylhexyl) phthalate (DEHP) metabolites (MECPP, MEHHP, MEOHP, and MEHP), especially in boys [@ShoaffCoullWeuve:2020]. Our findings, indicating that short-term childhood exposure to certain phthalate metabolites (MEHP, oh-MiNP, and oxo-MiNP) was associated with attentional function, adds to this growing evidence base suggesting that childhood phthalate exposure may impact child neurodevelopment.

Among phenols, some studies provide preliminary evidence of an association between BPA and {{< acr adhd >}} in children aged 8 to 15 years [@TewarAuingerBraun:2016] and in a case-control study of children aged 6 to 12 years [@LiZhangKuang:2018], especially in boys. Except for working memory, there was no evidence of an association between BPA and cognitive abilities in Spanish boys aged 9 to 11 years [@Rodriguez-CarrilloMustielesPerez-Lobato:2019]. We did not observe an association between BPA and attention function in the present study, but this study is the first to suggest that childhood exposure to MEPA may be associated with lower attentional function.

@@

We are also not aware of prior epidemiological studies specifically investigating the effects of elevated levels of glucocorticosteroids in relation to attentional function, although there is evidence that under- or over-production of glucocorticosteroids interfere with the normal development of the brain, possibly in a sex-specific manner [@LupienMcEwenGunnar:2009]. Our findings support evidence for such sex-specific interference.

Our findings should be interpreted considering the following strengths and limitations. Strengths include its relatively large sample size and its inclusion of multiple classes of non-persistent {{< acr edcs >}}. Further, this study used pooled urine samples for chemical assessment to reduce measurement error, since it is known that these specific {{< acr edcs >}} have very short half-lives [@PerrierGiorgis-AllemandSlama:2016; @CasasBasaganaSakhi:2018]. We decided to model both the *treatment* mechanisms, for the estimation of balancing weights, and the outcomes, with traditional covariates adjustment, to obtain *doubly robust* effect estimates. Finally, we decided not to interpret our results by focusing on the estimated coefficients of possibly misspecified regression models, but by making use of the g-computation procedure.

Limitations include the cross-sectional design of the present study. Importantly, the non-persistent {{< acr edcs >}} were measured in a pool of night and morning urine samples before the clinical visit, to represent exposure over the previous day, whereas the glucocorticosteroids were measured in the night urine sample. Cross-sectional designs also have limitations to study the etiologic period of susceptibility of attentional function. Prior studies have investigated associations between short-term effects of air pollution and attentional function, with null results in adolescents but higher response times in adults [@GignacBarrera-GomezPersavento:2021; @GignacRighiToran:2022]. Finally, longitudinal studies are necessary for a formal causal mediation analysis, which is why we did not attempt such a analysis even for the sex-stratified case, where glucocorticosteroids were associated with {{< acr hitrtse >}}.
Although we included a wide range of confounders, there is the possibility, as with other observational studies, of residual confounding, which might lead to a bias away from the null. We decided not to include prenatal levels of the {{< acr edcs >}} as confounders due to the *small* correlation between prenatal and childhood levels of these non-persistent chemicals [@HaugSakhiCequier:2018].
There is further the possibility of misspecification of the outcome model, although we included a spline of the exposure to relax some of the linearity assumptions. Finally, we acknowledge that the present study does not consider the possible mixture effect of these chemicals. While methods for causal mixture analysis are available [@KeilBuckleyO:2020], we also emphasize that using these tools without thoughtful consideration may lead to misleading results [@WebsterWeisskopf:2020].
